---
title: 'First Draft Visualization Project'
author: 'Hernan Figueroa - Yufei Wu'
output: html_document
---

### Introduction
Our project explores electricity consumption and electricity prices in different areas of New York City. We start exploring seasonality at the monthly and then weekly levels. We then compare prices vs consumption. The electric grid system in New York state is complex and the prices are not only dependend on consumption in New York City but also on prices from surrounding areas. 

### Analysis of data quality

The data were obtain from the New York Independent System Operator (NYISO):
- [New York Independent System Operator](https://www.nyiso.com/)
NOTE: The ISO website does not allow a bulk download of more than 150,000 records. The dataset for each year has to be downloaded separately

The datasets are time series for consumption (Load) and price (LMBP). These values are supposed to be recorded every 5 minutes; however, sometimes the sensor miss to record a value or record a value at a point different from 5 minutes.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 10)
```
```{r libraries, echo=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```


### **V1 - Monthly Average Consumption in NYC in 2018**
After exploring the datasets, we only read the time and load column. We can see that most data falls within the expected time stamps.
reading the dataset, we change the column Higher in the summer months due to AC usage. 
```{r 1v1}

read.NYISO <- function(filename){
  NYCLoad_df <- read.csv(filename, 
                        colClasses =c("character",
                                      rep("NULL",2),
                                      "character"),
                        stringsAsFactors = FALSE)
  colnames(NYCLoad_df) <- c("Time","Load")
  NYCLoad_df$Datetime <- strptime(NYCLoad_df$Time, "%Y/%m/%d %H:%M:%S")
  NYCLoad_df <- NYCLoad_df[1:(nrow(NYCLoad_df)-1),] #Last values belongs to next month
  NYCLoad_df$YMonth <- as.factor(format(NYCLoad_df$Datetime,"%Y-%m"))
  NYCLoad_df$Load <- as.numeric(NYCLoad_df$Load)
  return(NYCLoad_df)
}
NYCLoad_2018 <- read.NYISO('data/2018_NYC_Nov_OASIS_Real_Time_Dispatch_Actual_Load.csv')
Load_mean_2018 <-  na.omit(NYCLoad_2018[,c("YMonth","Load")]) %>%
                      group_by(YMonth) %>%
                      summarise(AvgLoad = mean(Load))
gv1 <- ggplot(Load_mean_2018, aes(YMonth,AvgLoad)) +
  geom_col() +
  ylab("")
gv1

```

### **V2 - Histogram of Consumption in NYC in 2018**
The summer months show some high values and more spreaded distributions. It may be more informative than V1
```{r 1v2}
gv2 <- ggplot(data = na.omit(NYCLoad_2018[,c("YMonth","Load")]), aes(Load)) +
  geom_histogram(fill = "red", binwidth = 100) +
  facet_wrap(~YMonth)
gv2
```

### **Is there seasonality in electricity consumption?**

NOTE: The ISO website does not allow a bulk download of more than 150,000 records. Each year has to be downloaded separately
```{r 1v3}
NYCLoad_2015 <- read.NYISO('data/2015_NYC_Nov_OASIS_Real_Time_Dispatch_Actual_Load.csv')
Load_mean_2015 <-  na.omit(NYCLoad_2015[,c("YMonth","Load")]) %>%
                      group_by(YMonth) %>%
                      summarise(AvgLoad = mean(Load))

NYCLoad_2016 <- read.NYISO('data/2016_NYC_Nov_OASIS_Real_Time_Dispatch_Actual_Load.csv')
Load_mean_2016 <-  na.omit(NYCLoad_2016[,c("YMonth","Load")]) %>%
                      group_by(YMonth) %>%
                      summarise(AvgLoad = mean(Load))

NYCLoad_2017 <- read.NYISO('data/2017_NYC_Nov_OASIS_Real_Time_Dispatch_Actual_Load.csv')
Load_mean_2017 <-  na.omit(NYCLoad_2017[,c("YMonth","Load")]) %>%
                      group_by(YMonth) %>%
                      summarise(AvgLoad = mean(Load))

levels(Load_mean_2015$YMonth) <- month.abb[c(11:12,1:10)]
levels(Load_mean_2016$YMonth) <- month.abb[c(11:12,1:10)]
levels(Load_mean_2017$YMonth) <- month.abb[c(11:12,1:10)]
levels(Load_mean_2018$YMonth) <- month.abb[c(11:12,1:10)]

Load_mean_2015$Year <- "2015"
Load_mean_2016$Year <- "2016"
Load_mean_2017$Year <- "2017"
Load_mean_2018$Year <- "2018"
Load_mean_df <- rbind(Load_mean_2015,Load_mean_2016,Load_mean_2017,Load_mean_2018)

gv3 <- ggplot(Load_mean_df, aes(YMonth, AvgLoad, color=Year)) + geom_point(size=4) + 
    ggtitle("Average Electricity Consumption in NYC") +
    labs (x = "", y = "MWh") +
    theme_grey(16) +
    theme(legend.title = element_blank())
gv3
```